---
title: "Data Cleaning"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Data Cleaning

Here's where we load the 3' tagseq samples, normalize them, and assemble them into a single count matrix.

Working with RNAseq data published in Krieger et al. 2020

```{r package-loading, cache=TRUE, echo = FALSE}
sapply(c("dplyr", "readr", "tidyr", "purrr", "ggplot2", "ggpubr", "openxlsx", "matrixStats"), require, character.only=TRUE)
```

## Reading in tagseq data and separating allele-specific counts

Reading in quantified RNAseq counts as individual *.ReadsPerGene.out.tab files, one file per sample. Counts for each allele of each gene are on separate rows. Systematic gene names are prepended with the species name (example: cer_YGR192C is the TDH3 ortholog from S. cerevisiae). Logistically, it'll work better for the analysis if we condense this format into only one row per gene (example: one YGR192C row). To do this, we first have to separate the counts for the Scer ortholog from the counts for the Spar ortholog:

```{r, cache=TRUE, dependson="package-loading"}
tagseq <- list.files("data_files/tagseq_counts/", 
                     full.names = TRUE) |> 
  map(read_table, col_names = FALSE, show_col_types = FALSE) |> 
  map(.f = select, X1, X3) |> # X1 are gene names, X3 is the sense strand read count
  purrr::reduce(.f = \(x, y) full_join(x = x, y = y, by = "X1"))

colnames(tagseq) <- c("gene", gsub("_ReadsPerGene.out.tab", "", list.files("data_files/tagseq_counts/", full.names = FALSE)))
QCdf <- tagseq[grepl("N_", tagseq$gene), ]
tagseq <- tagseq[!grepl("N_", tagseq$gene),]
tagseq <- tagseq[!tagseq$gene %in% c("cer_NA", "par_NA"),]
tagseq_cer <- tagseq[grepl("^cer_", tagseq$gene),]
tagseq_par <- tagseq[grepl("^par_", tagseq$gene),]
common_genes <- intersect(gsub("^cer_", "", tagseq_cer$gene),
                          gsub("^par_", "", tagseq_par$gene))
tagseq_cer$gene <- gsub("^cer_", "", tagseq_cer$gene)
tagseq_par$gene <- gsub("^par_", "", tagseq_par$gene)
tagseq_cer <- tagseq_cer[sapply(common_genes, \(x) which(x == tagseq_cer$gene)),]
rownames(tagseq_cer) <- common_genes
tagseq_par <- tagseq_par[sapply(common_genes, \(x) which(x == tagseq_par$gene)),]
rownames(tagseq_par) <- common_genes
sum(tagseq_cer$gene == tagseq_par$gene)
length(common_genes)
```

## Calculating % mapping to each allele for parents vs hybrids

In the parental samples, this means limiting to those reads that mapped to the "correct" parent's allele: in Scer samples, we use Scer allele counts, in Spar samples, we use Spar allele counts. In the hybrid this means splitting allele reads into separate columns.

But before we do this, we should make sure there aren't a lot of reads mapping to the "wrong" parent

Before we check percent mapping, we should filter out genes with very few reads total, as they'll have highly variable percents based only on a few reads.

```{r}
# normalizing function for filtering out lowly expressed 
# genes prior to assessing mapping bias
# (used later to actually normalize count data)
# normalizing counts to adjust for differences in library size
# sums .cts_cer and .cts_par to get library size, only returns
# counts for specified allele
# @input: count matrix (genes are rows, columns are samples)
# @output: a count matrix normalzied for library size---integer counts in counts-per-million
countsPerMillionAllele <- function(.cts_cer, .cts_par, .allele) {
  librarySizes <- colSums(.cts_cer, na.rm = TRUE) + colSums(.cts_par, na.rm = TRUE)
  if (.allele == "cer") {
    .cts <- .cts_cer
  }
  if (.allele == "par") {
    .cts <- .cts_par
  }
  output <- apply(.cts, 1, function(x) {
    normalized <- (x/librarySizes)*1e6
    return(round(normalized))
  })
  return(t(output)) # For some unhinged reason, a vector output of apply across ROWS forms the COLUMNS of a new matrix
}
# tests for countsPerMillionAllele
test_cts <- tagseq_cer[,-1]
test_rowIdx <- sample(c(1:nrow(test_cts)), 1)
test_colIdx <- sample(which(grepl("cer", colnames(test_cts))), 1)
test_count <- test_cts[test_rowIdx, test_colIdx]
test_cpm <- countsPerMillionAllele(.cts_cer = tagseq_cer[,-1], 
                                   .cts_par = tagseq_par[,-1],
                                   .allele = "cer")
((test_count/(colSums(tagseq_cer[,-1], na.rm = TRUE) + 
                colSums(tagseq_par[,-1], na.rm = TRUE))[test_colIdx])*1e6) %>% 
  round() # what it should be
test_cts[test_rowIdx, test_colIdx] # what it is before using our function
test_cpm[test_rowIdx, test_colIdx] # what it is using our function
```
Now we can calculate percent reads mapping to each allele in parental samples

```{r}
# 1) normalize to counts per million based on total 
# library size: cer reads + par reads regardless of sample organism
cpm_cer <- countsPerMillionAllele(.cts_cer = tagseq_cer[,-1],
                                      .cts_par = tagseq_par[,-1],
                                      .allele = "cer")
cpm_par <- countsPerMillionAllele(.cts_cer = tagseq_cer[,-1],
                                      .cts_par = tagseq_par[,-1],
                                      .allele = "par")
# 2) filter lowly expressed: < 30 cpm
sum(cpm_cer == 0 & cpm_par == 0)
isHighExpr <- (rowMeans(cpm_cer + cpm_par) > 30) |> sapply(FUN = isTRUE)
keep_genes <- common_genes[isHighExpr]
cpm_cer <- cpm_cer[isHighExpr,]
cpm_par <- cpm_par[isHighExpr,]
sum(cpm_cer == 0 & cpm_par == 0) # note there are still individual samples with zero counts

# 3) check % cer of all high-enough expressed genes is close to 1 for cer samples and 0 for par samples
plotdf <- bind_rows(bind_cols(tibble(gene = keep_genes,
                                     allele = "cer"), cpm_cer),
                    bind_cols(tibble(gene = keep_genes,
                                     allele = "par"), cpm_par)) |> 
  pivot_longer(cols = colnames(tagseq_cer[,-c(1,2)]),
               names_to = c("sample_name"),
               values_to = "count") |> 
  pivot_wider(id_cols = c("sample_name", "gene"),
              values_from = "count", names_from = "allele",
              names_prefix = "counts_")
plotdf$organism <- if_else(grepl("_cer_", plotdf$sample_name),
                           true = "cerSample", 
                           false = if_else(grepl("_par_", plotdf$sample_name),
                                           true = "parSample",
                                           false = "hybSample"))
# Calculating % of reads mapping to the Scer allele 
# for each gene/sample
# (So % Spar is 1 - % Scer)
plotdf$pct_cer <- if_else(plotdf$counts_cer == 0 &
                            plotdf$counts_par == 0,
                          true = NA,
                          false = plotdf$counts_cer/(plotdf$counts_cer + plotdf$counts_par))
plotdf <- drop_na(plotdf)
sample_genes <- sample(plotdf$gene, size = 100)
cer_genedf <- plotdf |> filter(organism == "cerSample") |> 
  group_by(gene) |> 
  summarise(avg_pct_cer = mean(pct_cer, na.rm = TRUE))
hist(cer_genedf$avg_pct_cer, breaks = 50)
abline(v = 0.9, col = "red")
par_genedf <- plotdf |> filter(organism == "parSample") |> 
  group_by(gene) |> 
  summarise(avg_pct_cer = mean(pct_cer, na.rm = TRUE))
hist(par_genedf$avg_pct_cer, breaks = 50)
abline(v = 0.1, col = "red")
```

As expected, genes from Scer samples have a high percentage of reads mapping to the Scer gene ortholog and vice versa for Spar.

## Combining allele-specific counts

Now we are ready to combine allele-specific counts into a single count matrix. To do this, we need to know which samples come from which organism, so we will first load our sample metadata.

```{r}
# tagseq
# reading in sample info
sample_info <- read.xlsx("data_files/sample_metadata/bioSample1to999.xlsx", na.strings="not applicable", cols=c(1,4,9,13,14,15,17)) %>%
  bind_rows(read.xlsx("data_files/sample_metadata/bioSample1000toEnd.xlsx", na.strings="not applicable", cols=c(1,4,9,13,14,15,17)))
colnames(sample_info) <- c("sample_name", "organism" , "collection_date", "genotype", "experiment","time_point", "well_flask_ID")
# removing samples not from LowN experiment
sample_info <- filter(sample_info, experiment == "YPD to Low N")
cat("All metadata samples are in counts matrix Scer:", all(sample_info$sample_name %in% colnames(tagseq_cer)), "\n")
cat("All metadata samples are in counts matrix Spar:", all(sample_info$sample_name %in% colnames(tagseq_par)), "\n")
# creating count matrix
counts <- apply(sample_info, 1, \(x) {
  sample_name <- x["sample_name"]
  org <- x["organism"]
  if (!sample_name %in% colnames(tagseq_cer)) {
    cat("missing sample", sample_name, "\n")
    output <- matrix(NA, nrow = nrow(tagseq_cer), ncol = 1)
    colnames(output) <- sample_name
    return(output)
  }
  if (org == "Saccharomyces cerevisiae") {
    return(tagseq_cer[,sample_name, drop = FALSE])
  }
  if (org == "Saccharomyces paradoxus") {
    return(tagseq_par[,sample_name, drop = FALSE])
  }
  if (org == "Saccharomyces cerevisiae x Saccharomyces paradoxus") {
    cer_countcol <- tagseq_cer[,sample_name]
    par_countcol <- tagseq_par[,sample_name]
    hyb_countcol <- tagseq_cer[,sample_name] + tagseq_par[,sample_name]
    output <- cbind(cer_countcol, par_countcol, hyb_countcol)
    colnames(output) <- c(gsub("_hyb_", "_hyc_", sample_name),
                          gsub("_hyb_", "_hyp_", sample_name), sample_name)
    return(output)
  }
}) |> Reduce(f = cbind)
rownames(counts) <- common_genes
cat("All gene names are the same in combined count matrix and Scer count matrix:", all(rownames(counts) == rownames(tagseq_cer)), "
    \n")
cat("All gene names are the same in combined count matrix and Spar count matrix:", all(rownames(counts) == rownames(tagseq_par)), "
    \n")
# adding two more rows for each hybrid allele in info df
sample_info <- map(c(1:nrow(sample_info)), \(i) {
  x <- sample_info[i,]
  org <- sample_info[i,"organism"]
  if (org == "Saccharomyces cerevisiae x Saccharomyces paradoxus") {
    x_cer <- x
    x_par <- x
    x_cer["sample_name"] <- gsub("_hyb_", "_hyc_", x_cer["sample_name"])
    x_par["sample_name"] <- gsub("_hyb_", "_hyp_", x_par["sample_name"])
    output <- bind_rows(x_cer, x_par, x)
    return(output)
  }
  if (org != "Saccharomyces cerevisiae x Saccharomyces paradoxus") {
    return(x)
  }
}) |> purrr::reduce(.f = bind_rows)
cat("All samples in the same order in count matrix and metadata:", all(colnames(counts) == sample_info$sample_name), "\n")
```
## cleaning up some sample info column values

```{r}
# shorten organism names
sample_info$organism <- map_chr(sample_info$sample_name, function(s) {
  if (grepl("_cer_", s)) {
    return("cer")
  }
  if (grepl("_par_", s)) {
    return("par")
  }
  if (grepl("_hy[bpc]_", s)) {
    return("hyb")
  }
})

# add allele column
sample_info$allele <- map_chr(sample_info$sample_name, function(s) {
  if (grepl("_cer_", s) | grepl("_hyc_", s)) {
    return("cer")
  }
  if (grepl("_par_", s) | grepl("_hyp_", s)) {
    return("par")
  }
  if (grepl("_hyb_", s)) {
    return("hyb")
  }
})

# removing space from genotype
sample_info$genotype <- gsub(" ", "", sample_info$genotype)

# converting timepoint to integer values of minutes
timepoint_to_int <- function(t) {
  if (grepl("[0-9] h", t)) {
    return(parse_number(t)*60)
  }
  else {
    return(parse_number(t))
  }
}
sample_info$time_point_num <- map_dbl(sample_info$time_point, timepoint_to_int)
colnames(sample_info) <- map_chr(colnames(sample_info), gsub, pattern = "^time_point$", replacement = "time_point_str") # time_point_str is the version that we'll use for DESeq2, so we can set a reference level (but we'll have to do that later)

# preserving sample information that is non-unique for replicates
sample_info$condition <- paste(sample_info$genotype,
                               sample_info$time_point_num, sep="_")

# also creating a condition variable that includes organism (for determining replicate samples)
sample_info$org_gen_tp <- paste(sample_info$organism,
                                sample_info$genotype,
                                sample_info$time_point_num, sep="_")

# In case you're wondering, the collection date is when RNA was collected, 
# NOT when the living yeast sample was collected
# (only like 65 samples total have a collection date within 24 hours 
# for all 3 samples, and that's mostly because a lot of samples were collected on those dates)
sample_info <- select(sample_info, !collection_date)

# Check for missing (NA) values
geneHasNAs <- apply(counts, 1, function(x) {
  isNA <- sapply(x, is.na)
  return(any(isNA))
}) 
sum(geneHasNAs) # should have 0
```

## Renaming replicates

Issue: well_flask_IDs, which are supposed to be the same ID for all three samples taken from the same well at each of the three timepoints, are currently slightly different for the same well. The reason is that there is a unique sample tag at the end of the well_flask_ID that we need to remove.

Parent samples currently have mainly 1-2 wells labeled with the same well_flask_ID:

```{r, message=FALSE}
sample_info |> filter(organism != "hyb") |> 
  mutate(org_gen = paste(organism, genotype, sep = "_")) |>
  select(well_flask_ID, org_gen) |> 
  table() |> table() # (0s are combinations of organism/well_flask_ID/genotype that don't exist, which is common because many well flask IDs are only present in one organism or genotype)
```

And Hybrids currently have mainly 3-6:

```{r, message=FALSE}
sample_info |> filter(organism == "hyb") |> 
  select(well_flask_ID, genotype) |> 
  table() |> table()
```
By the end of this section, every parental sample should still have at most 3 entries and most should have 3 (9 for hybrids, one for each of the two alleles plus summed alleles).

First of all (based on personal communication), the GS2018 samples (which are the majority of the WT samples) have a different well_flask_ID for the 1 hr sample than for 0 or 16 hr. For each row in the well plate (A-H), these are the numbers that are paired: 1-7, 2-8, 3-9, 4-10, 5-11, 6-12:
```{r}
col1 <- sapply(c("A", "B", "C", "D", "E", "F", "G", "H"), function(x) return(paste0(x, c(1:6)))) %>% as.vector()
col2 <- sapply(c("A", "B", "C", "D", "E", "F", "G", "H"), function(x) return(paste0(x, c(7:12))))  %>% as.vector()
gs2018_lookup <- tibble(TP1_TP3 = col1, TP2 = col2)

# arbitrarily assigning each ID its TP1/TP3 (column 1 in lookup table) value as opposed to its TP2 value
standardizeGS2018ID <- function(id) {
  id_clipped <- gsub("_WT2_GS2018", "", id)
  id_clipped <- gsub("_G12_GS2018", "", id_clipped)
  id_clipped <- gsub("_G6_GS2018", "", id_clipped)
  rownum <- c(which(gs2018_lookup$TP1_TP3 == id_clipped), which(gs2018_lookup$TP2 == id_clipped))
  new_id <- gsub(gs2018_lookup$TP2[rownum], gs2018_lookup$TP1_TP3[rownum], id)
  return(new_id)
}
# tests for standardizeGS2018ID
standardizeGS2018ID("G8_G12_GS2018")
```

Before applying:
```{r}
cat("Parent samples (all 1s and 2s):\n")
sample_info |> filter(organism %in% c("cer", "par") & grepl("GS2018", well_flask_ID)) |> 
  select(well_flask_ID) |> table() 
cat("Hybrid samples (all 3s and 6s):\n")
sample_info |> filter(organism == "hyb" & grepl("GS2018", well_flask_ID)) |> 
  select(well_flask_ID) |> table()
```

Applying to GS2018 samples:
```{r}
GS2018_idxs <- grepl("GS2018", sample_info$well_flask_ID)
sample_info$well_flask_ID[GS2018_idxs] <- sapply(sample_info$well_flask_ID[GS2018_idxs], standardizeGS2018ID)
```

After applying:
```{r}
cat("Parent samples (should have 3, except for H1, which is missing a timepoint):\n")
sample_info |> filter(organism %in% c("cer", "par") & grepl("GS2018", well_flask_ID)) |> 
  select(well_flask_ID) |> table() 
cat("Hybrid samples (should have 9):\n")
sample_info |> filter(organism == "hyb" & grepl("GS2018", well_flask_ID)) |> 
  select(well_flask_ID) |> table()
# hybrids should have 9, (hyc, hyp, and hyb) x 3 timepoints, Parents have just 3 timepoints. Except for H1, which is missing a timepoint
```
Now we can get rid of the "GS" part of the rep name, which can be different for the same sample at different timepoints and replacing it with the tag immediately before the well_flask_ID in the sample name for non-unique well_flask_IDs
```{r}
sample_info$well_flask_ID <- gsub("_GS.*", "", sample_info$well_flask_ID)
sample_info$new_id <- map2(sample_info$well_flask_ID, 
             sample_info$sample_name, \(i, s) {
               ex <- sample_info |> filter(sample_name == s) |> 
                 select(experiment) |> pull()
               org <- sample_info |> filter(sample_name == s) |> 
                   select(organism) |> pull()
                 is_duplicate <- sample_info |> 
                   filter(organism == org &
                            well_flask_ID == i) |> 
                   group_by(time_point_str) |> 
                   summarise(n_per_tp = n())
                 if (any(is_duplicate$n_per_tp > 1)) {
                   s <- gsub("_GS.*", "", s)
                   new_tag <- gsub(i, "", s) |> strsplit(split = "_") |> 
                     unlist() |> tail(n = 1)
                   return(paste(new_tag, i, sep = "_"))
                 }
                 else {
                   return(i)
                 }
             }) |> unlist()
# checking for non-unique IDs
sample_info |> 
  group_by(new_id, organism, genotype, time_point_str) |> 
  summarise(n_per_org_gen_tp = n()) |> 
  filter((organism != "hyb" & n_per_org_gen_tp > 1) |
           (organism == "hyb" & n_per_org_gen_tp > 2)) # should be empty
```

```{r}
# updating well_flask_ID 
sample_info$well_flask_ID <- sample_info$new_id
sample_info <- select(sample_info, -"new_id")

# checking example (two C5_A10s for the 960 timepoint)
sample_info |> filter(organism == "par" &
                        genotype == "GCN4delete") |> 
  select(sample_name, condition, well_flask_ID)
# should have additional tag P1 or P2 on the C5_A10s
```

Final check that now for parental samples: a) most well flask IDs are represented 3 times, and b) no well flask ID is represented more than 3 times for each organism/condition:

```{r, message=FALSE}
sample_info |> filter(organism != "hyb") |> 
  mutate(org_gen = paste(organism, genotype, sep = "_")) |>
  select(well_flask_ID, org_gen) |> 
  table() |> table() # 0s are combinations of organism/well_flask_ID/genotype that don't exist, which is common because many well flask IDs are only present in one organism or genotype
```

And for hybrid samples: a) most well flask IDs are represented 9 times, and b) no well flask ID is represented more than 9 times for each organism/condition:

```{r, message=FALSE}
sample_info |> filter(organism == "hyb") |> 
  select(well_flask_ID, genotype) |> 
  table() |> table()
```

## Removing samples with small library sizes

```{r}
# Exploring library sizes (un-normalized)
libsizes <- colSums(counts) # these counts include all samples, hybrid and parental
ggplot(tibble(libsize = libsizes), aes(x = libsize)) + geom_histogram()
sort(libsizes)[c(1:20)]
```

No libraries under 100k reads, no need to remove small libraries prior to normalizing.

Are the hybrid library sizes correlated between hyc and hyp alleles?

```{r}
plotdf <- tibble(sample_name = names(libsizes[grepl("_hy[pc]", names(libsizes))]),
                 libsize = libsizes[grepl("_hy[pc]", names(libsizes))]) |> 
  left_join(y = select(filter(sample_info, organism == "hyb"), sample_name, experiment), by = "sample_name")
plotdf$allele <- if_else(grepl(pattern = "_hyc_", plotdf$sample_name),
                         true = "cer", false = "par")
plotdf$sample_name <- gsub("_hy[pc]_", "_hyb_", plotdf$sample_name)
plotdf <- plotdf |> pivot_wider(id_cols = c("sample_name", "experiment"), 
                                names_from = allele,
                                values_from = libsize)
ggplot(plotdf, aes(x = cer, y = par)) + 
  geom_point(aes(color = experiment), alpha = 0.5) +
  geom_text(data = filter(plotdf, cer > par*1.5 | par > cer*1.5),
            aes(label = sample_name), check_overlap = TRUE, color = "green") +
  geom_abline(intercept = 0, slope = 1, color = "red") +
  geom_rect(xmin = 0, xmax = 100000, ymin = 0, ymax = 100000, color = "blue", alpha = 0) +
  geom_vline(xintercept = 100000, color = "blue", alpha = 0.5) +
  geom_hline(yintercept = 100000, color = "blue", alpha = 0.5)
# (the samples within the blue box will be removed by our library size threshold below)
# and no other samples should be below the blue lines but not within the blue box
```

Conclusion: yes, very correlated.

## Normalizing to counts per million

In tagseq data, 1 read = one mRNA molecule. So normalizing to counts per million is simply dividing the count by the library size * 1 million.

```{r}
# normalizing counts to adjust for differences in library size
# @input: count matrix (genes are rows, columns are samples)
# @output: a count matrix normalzied for library size---integer counts in counts-per-million
countsPerMillion <- function(.cts) {
  librarySizes <- colSums(.cts, na.rm = TRUE)
  output <- apply(.cts, 1, function(x) {
    normalized <- (x/librarySizes)*1e6
    return(round(normalized))
  })
  return(t(output)) # For some unhinged reason, a vector output of apply across ROWS forms the COLUMNS of a new matrix
}
# tests for countsPerMillion
test_cts <- counts[,grepl("_par", colnames(counts))]
test_rowIdx <- sample(c(1:nrow(test_cts)), 1)
test_colIdx <- sample(c(1:ncol(test_cts)), 1)
test_count <- test_cts[test_rowIdx, test_colIdx]
test_cpm <- countsPerMillion(test_cts)
test_cpm_cpm <- countsPerMillion(test_cpm)
((test_count/colSums(test_cts, na.rm = TRUE)[test_colIdx])*1e6) %>% round() # what it should be
test_cts[test_rowIdx, test_colIdx] # what it is before using our function
test_cpm[test_rowIdx, test_colIdx] # what it is using our function
test_cpm_cpm[test_rowIdx, test_colIdx] # what it is if you run cpm too many times (should be the same as test_cpm)
```

Actually normalizing

```{r}
counts_unnorm <- counts
counts <- countsPerMillion(counts)
```

## Saving

```{r}
save(counts, counts_unnorm, sample_info, file = "data_files/CleanedCounts.RData")
```

