---
title: "Fig_TFdel"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, warning=FALSE}
sapply(c("dplyr", "readr", "tidyr", "purrr", "ggplot2", "ggpubr", "matrixStats", "ComplexHeatmap", "circlize", "ggVennDiagram", "ggExtra"), require, character.only=TRUE)
load("data_files/CleanedCounts.RData")
load("data_files/DESeq2.RData")
load("data_files/GeneDataFrame.RData")
source("utils.R")

sgd_lookup <- read_delim("data_files/downloaded_genomes_and_features/SGD_features.tab", delim = "\t", col_names = FALSE) |> filter(X2 == "ORF") |> 
  select(X4, X5) |> 
  dplyr::rename("gene_name"="X4",
                "common_name"="X5")

FullTFnames <- setdiff(unique(gsub("delete", "", sample_info$genotype)), "WT")
NitrogenTFs <- c("URE2", "GAT1", "GLN3", "DAL80", "GZF3")
# ordering TFnames so that nitrogen comes first then alphabetical
TFnames <- c(NitrogenTFs, sort(setdiff(FullTFnames, NitrogenTFs)))

p_thresh <- 0.05 # because DESeq2 already corrected for FDR with alpha = 0.05
eff_thresh <- 1
TFdel_lookup <- read_delim("data_files/downloaded_genomes_and_features/yeastract_46TFs.csv", col_names = FALSE, col_select = c(1,2), delim = ";") # gets some warnings, but so far has been fine
colnames(TFdel_lookup) <- c("common", "systematic")
```

## Which TFs are differentially expressed between species?

Even though this is only using WT data, this is more relevant to the TFdels, so we'll start with this. Mean expression in Scer vs Spar, averaged across timepoints and replicates. Error bars are standard error from replicates.

```{r}
# scatter plot Scer vs Spar with geom_text as the TF names (b/c the heatmap was only useful when there was more than one environment)
plotdf <- tibble(tf = TFnames) |> 
  left_join(y = sgd_lookup, 
            by = c("tf"="common_name"))
plotdf$mean_expr_cer <- map(plotdf$gene_name, \(g) {
  expr_vec <- NA
  if (g %in% rownames(counts)) {
    expr_vec <- counts[g, sample_info$genotype == "WT" &
                       sample_info$organism == "cer"] |> 
    as.numeric()
  }
  return(mean(expr_vec))
}) |> unlist()
plotdf$mean_expr_par <- map(plotdf$gene_name, \(g) {
  expr_vec <- NA
  if (g %in% rownames(counts)) {
    expr_vec <- counts[g, sample_info$genotype == "WT" &
                       sample_info$organism == "par"] |> 
    as.numeric()
  }
  return(mean(expr_vec))
}) |> unlist()

print(ggplot(plotdf, aes(x = log2(mean_expr_cer + 1), 
                   y = log2(mean_expr_par + 1))) +
  geom_text(aes(label = tf, color = tf %in% NitrogenTFs)) +
  geom_abline(slope = 1, intercept = 0))
```

## How many genes does each TFdel affect in each species?

How many genes are affected in shams vs real knockout lines?
```{r}
plotdf <- bind_rows(mutate(TFdeldf, type = "real"),
                    mutate(TFdeldf_sham1, type = "sham1"),
                    mutate(TFdeldf_sham2, type = "sham2")) |> 
  filter(organism == allele) |> 
  select(gene_name, organism, lfc, padj, type) |> 
  group_by(organism, type) |> 
  summarise(nSig = sum(abs(lfc) > eff_thresh & padj < p_thresh))
p <- ggplot(plotdf, aes(x = type, 
                        y = nSig)) +
  geom_histogram(aes(fill = organism), stat = "identity") +
  geom_text(aes(x = type, y = nSig, label = nSig), nudge_y = 500) +
  scale_fill_discrete(limits = c("cer", "par", "hyb"),
                      type = c(conservation_colordf$value[conservation_colordf$limits == "Scer"],
                               conservation_colordf$value[conservation_colordf$limits == "Spar"],
                               conservation_colordf$value[conservation_colordf$limits == "Hyb"])) +
  facet_wrap(~factor(organism, levels = c("cer", "par", "hyb"),
                     labels = c("Scer", "Spar", "Hyb"))) +
  theme_classic() +
  xlab("") +
  ylab("Number of significantly differentiall expressed\ngenes across all 46 TF knockouts") +
  theme(legend.position = "none")
p
```

For the paper figure:
```{r figure_sham}
# pdf("../Umich_Dissertation/chapters/figures/Networks/shams.pdf",
#     width = 5, height = 4)
# p
# dev.off()
```


Barplots of nGenes up or down in response to each TF deletion.

If we only consider a TF effect shared if it affects the same gene in multiple yeasts and occurs at the same timepoint and direction:
```{r timepoint-specific-effects}
plotdf <- TFdeldf |> filter(padj < p_thresh &
                             abs(lfc) > eff_thresh) |> 
  mutate(lfc_sign = sign(lfc)) |> 
  select(gene_name, deletion, organism, timepoint, lfc_sign) |>
  group_by(gene_name, deletion, timepoint, lfc_sign) |> 
  summarise(orgs = list(organism))
cat("Fraction of single-yeast effects:", sum(sapply(plotdf$orgs, length) == 1), "/", nrow(plotdf),
    "(", sum(sapply(plotdf$orgs, length) == 1)/nrow(plotdf), "%)", "\n")
```
If we instead consider a TF effect shared if it affects the same gene in multiple yeasts in the same direciton, but occurs at any timepoint:
```{r timepoint-agnostic-effects}
plotdf <- TFdeldf |> filter(padj < p_thresh &
                             abs(lfc) > eff_thresh) |> 
  mutate(lfc_sign = sign(lfc)) |> 
  select(gene_name, deletion, organism, lfc_sign) |>
  #unique() |> # unique stops double-counting of gene-deletion-lfc_sign combos that affect multiple timepoints, but that currently results in single-species effects becoming a larger portion of the total effects than the timepoint-specific chunk above, which is counterintuitive so commenting it for now
  group_by(gene_name, deletion, lfc_sign) |> 
  summarise(orgs = list(organism))
cat("Fraction of single-yeast effects:", sum(sapply(plotdf$orgs, length) == 1), "/", nrow(plotdf),
    "(", sum(sapply(plotdf$orgs, length) == 1)/nrow(plotdf), "%)", "\n")
```
That's not a very big difference, so for now we'll specify they're shared if they affect the same TF. Venn diagram makes it even clearer that it doesn't matter so we might as well choose the loosest, simplest definition:

```{r venn}
sigdf <- filter(TFdeldf, padj < p_thresh & abs(lfc) > eff_thresh & organism == allele)
plotdf <- sigdf |> filter(organism == allele)
# TF-gene combo is shared:
p <- ggVennDiagram(list(Scer = paste(plotdf$gene_name[plotdf$organism == "cer"], 
                                plotdf$deletion[plotdf$organism == "cer"], sep = "_"), 
                   Spar = paste(plotdf$gene_name[plotdf$organism == "par"], 
                                plotdf$deletion[plotdf$organism == "par"], sep = "_"),
                   Hyb = paste(plotdf$gene_name[plotdf$organism == "hyb"], 
                                plotdf$deletion[plotdf$organism == "hyb"], sep = "_"))) + 
                scale_fill_gradient(low="grey90",high = "red") +
  ggtitle("TF-gene combination is shared")
# TF-gene-direction combo is shared:
p_dir <- ggVennDiagram(list(Scer = paste(plotdf$gene_name[plotdf$organism == "cer"], 
                                plotdf$deletion[plotdf$organism == "cer"],
                                sign(plotdf$lfc)[plotdf$organism == "cer"], sep = "_"), 
                   Spar = paste(plotdf$gene_name[plotdf$organism == "par"], 
                                plotdf$deletion[plotdf$organism == "par"],
                                sign(plotdf$lfc)[plotdf$organism == "par"], sep = "_"),
                   Hyb = paste(plotdf$gene_name[plotdf$organism == "hyb"], 
                               plotdf$deletion[plotdf$organism == "hyb"],
                               sign(plotdf$lfc)[plotdf$organism == "hyb"], sep = "_"))) + 
  scale_fill_gradient(low="grey90",high = "red") +
  ggtitle("TF-gene-direction combination is shared")
# TF-gene-direction-timepoint combo is shared:
p_dirTP <- ggVennDiagram(list(Scer = paste(plotdf$gene_name[plotdf$organism == "cer"], 
                                plotdf$deletion[plotdf$organism == "cer"],
                                sign(plotdf$lfc)[plotdf$organism == "cer"],
                                plotdf$timepoint[plotdf$organism == "cer"], sep = "_"), 
                   Spar = paste(plotdf$gene_name[plotdf$organism == "par"], 
                                plotdf$deletion[plotdf$organism == "par"],
                                sign(plotdf$lfc)[plotdf$organism == "par"],
                                plotdf$timepoint[plotdf$organism == "par"], sep = "_"),
                   Hyb = paste(plotdf$gene_name[plotdf$organism == "hyb"], 
                               plotdf$deletion[plotdf$organism == "hyb"],
                               sign(plotdf$lfc)[plotdf$organism == "hyb"],
                                plotdf$timepoint[plotdf$organism == "hyb"], sep = "_"))) + 
  scale_fill_gradient(low="grey90",high = "red") +
  ggtitle("TF-gene-direction-timepoint combination is shared")
```
For the paper figure:
```{r figure_Venntfdel}
# pdf("../Umich_Dissertation/chapters/figures/Networks/VennTFdel.pdf",
#     width = 4, height = 4)
# p
# dev.off()
```

Number of TFs each gene is affected by in each background:
```{r nTF histograms}
plotdf <- expand_grid(gene_name = genedf$gene_name,
                      organism = c("cer", "hyb", "par"))
sigdf <- TFdeldf |> filter(abs(lfc) > eff_thresh & padj < p_thresh)
plotdf$nTFs <- map2(plotdf$gene_name, plotdf$organism, \(g, o) {
  df_go <- sigdf |> filter(gene_name == g & organism == o & allele == o) |> 
    select(deletion) |> unique()
  return(nrow(df_go))
}) |> unlist()
p <- ggplot(plotdf, aes(x = nTFs)) +
  geom_histogram(stat = "count", aes(fill = organism)) +
  scale_fill_discrete(limits = c("cer", "par", "hyb"),
                      type = c(conservation_colordf$value[conservation_colordf$limits == "Scer"],
                               conservation_colordf$value[conservation_colordf$limits == "Spar"],
                               conservation_colordf$value[conservation_colordf$limits == "Hyb"])) +
  facet_wrap(~factor(organism, levels = c("cer", "par", "hyb"),
                     labels = c("Scer", "Spar", "Hyb"))) +
  theme_classic() +
  ylab("Number of genes") +
  xlab("Number of TF knockouts that significantly affect expression of each gene") +
  theme(legend.position = "none")
p
```

For the paper figure:
```{r figure_powerLaw}
# pdf("../Umich_Dissertation/chapters/figures/Networks/nTFs.pdf",
#     width = 9, height = 2)
# p
# dev.off()
```

## Allele-specific TF effects in the hybrid
```{r}
cat("correlation between Hyc and Hyp for all genes:")
plotdf <- TFdeldf |> filter(organism == "hyb" & allele != "hyb") |> 
  select(gene_name, deletion, timepoint, lfc, padj, allele) |> 
  pivot_wider(id_cols = c("gene_name", "deletion", "timepoint"),
              names_from = "allele", values_from = c("lfc", "padj")) 
cor(plotdf$lfc_cer, plotdf$lfc_par, use = "complete.obs")
cat("correlation when Hyc or Hyp are sig:")
plotdf_oneSig <- plotdf |> 
  filter(padj_cer < p_thresh | padj_par < p_thresh)
cor(plotdf_oneSig$lfc_cer, plotdf_oneSig$lfc_par, use = "complete.obs")
cat("correlation when both alleles are sig:")
plotdf_bothSig <- plotdf_oneSig |> 
  filter(padj_cer < p_thresh & padj_par < p_thresh)
cor(plotdf_bothSig$lfc_cer, plotdf_bothSig$lfc_par, use = "complete.obs")
cat("correlation when both alleles are sig and in the same direction:")
plotdf_bothSigSameDir <- plotdf_bothSig |> 
  filter(sign(lfc_cer) == sign(lfc_par))
cor(plotdf_bothSigSameDir$lfc_cer, plotdf_bothSigSameDir$lfc_par, use = "complete.obs")
cat("fraction of effects where both alleles are affected out of cases where at least one allele is affected:")
nrow(plotdf_bothSig)/nrow(plotdf_oneSig)
p <- ggplot(data = tibble(n = c(nrow(plotdf_bothSig), 
                           nrow(plotdf_oneSig) - nrow(plotdf_bothSig)),
                     type = c("both alleles", "one allele"))) + 
  geom_bar(aes(x = type, y = n), stat = "identity", 
           fill = "grey", color = "black") +
  xlab("") +
  ylab("number of significant expression effects\n across all 46 TF knockouts") +
  theme_classic() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)) +
  ggtitle("Hybrid allele-specific\n TF knockout effects")
p
```
For the paper figure
```{r hybrid barplot}
 pdf("../Umich_Dissertation/chapters/figures/Networks/hybridBarplot.pdf",
    width = 2.6, height = 3.5)
p
dev.off()
```


Fraction of times the combined Hyb is significant when one of the alleles is. (Just for curiousity on which type of count data is has more power in DESeq2)
```{r}
plotdf <- TFdeldf |> filter(organism == "hyb") |> 
  select(gene_name, deletion, timepoint, lfc, padj, allele) |> 
  pivot_wider(id_cols = c("gene_name", "deletion", "timepoint"),
              names_from = "allele", values_from = c("lfc", "padj")) |> 
  filter(padj_cer < p_thresh | padj_par < p_thresh)
sum(plotdf$padj_hyb < p_thresh)/nrow(plotdf)
```
Fraction of times the combined Hyb is significant when neither alleles is:
```{r}
plotdf <- TFdeldf |> filter(organism == "hyb") |> 
  select(gene_name, deletion, timepoint, lfc, padj, allele) |> 
  pivot_wider(id_cols = c("gene_name", "deletion", "timepoint"),
              names_from = "allele", values_from = c("lfc", "padj")) |> 
  filter(padj_cer >= p_thresh & padj_par >= p_thresh)
sum(plotdf$padj_hyb < p_thresh)/nrow(plotdf)
```

Example hybrid genes in each category
```{r example genes, results="hide"}
# first determine categories
plotdf <- TFdeldf |> filter(organism == "hyb" & allele != "hyb") |> 
  select(gene_name, deletion, timepoint, lfc, padj, allele) |> 
  pivot_wider(id_cols = c("gene_name", "deletion", "timepoint"),
              names_from = "allele", values_from = c("lfc", "padj")) |> 
  filter(padj_cer < p_thresh | padj_par < p_thresh)
# then select random genes from each group
highExpressedGenes <- rownames(counts)[rowMeans(counts) > 30]
gene_Hyc <- plotdf |> filter((padj_cer < p_thresh) & 
                               !(padj_par < p_thresh) &
                               (lfc_cer < -10) &
                               !(abs(lfc_par) > eff_thresh) &
                               gene_name %in% highExpressedGenes) |> 
  slice_sample(n = 1)
gene_Hyp <- plotdf |> filter(!(padj_cer < p_thresh) & 
                               (padj_par < p_thresh) &
                               !(abs(lfc_cer) > eff_thresh) &
                               (abs(lfc_par) > 2*eff_thresh) &
                               gene_name %in% highExpressedGenes) |> 
  slice_sample(n = 1)
gene_bothSame <- plotdf |> filter((padj_cer < p_thresh) & 
                                    (padj_par < p_thresh) &
                                    (abs(lfc_cer) > 2*eff_thresh) &
                                    (abs(lfc_par) > 2*eff_thresh) &
                                    (sign(lfc_par) == sign(lfc_cer)) &
                                    gene_name %in% highExpressedGenes) |> 
  slice_sample(n = 1)
gene_bothDifferent <- plotdf |> filter((padj_cer < p_thresh) & 
                                         (padj_par < p_thresh) &
                                         (abs(lfc_cer) > 2*eff_thresh) &
                                         (abs(lfc_par) > 2*eff_thresh) &
                                         (sign(lfc_par) != sign(lfc_cer)) &
                                         gene_name %in% highExpressedGenes) |> 
  slice_sample(n = 1)

p_Hyc <- plotGenesTFdel(.gene_idxs = gene_Hyc$gene_name, .tf = gene_Hyc$deletion,
                        .parents_or_hybrid = "hybrid") +
  ggtitle(paste("Hyc only\n", gene_Hyc$gene_name, gene_Hyc$deletion, gene_Hyc$timepoint))
p_Hyp <- plotGenesTFdel(.gene_idxs = gene_Hyp$gene_name, .tf = gene_Hyp$deletion,
                        .parents_or_hybrid = "hybrid") +
  ggtitle(paste("Hyp only\n", gene_Hyp$gene_name, gene_Hyp$deletion, gene_Hyp$timepoint))
p_bothSame <- plotGenesTFdel(.gene_idxs = gene_bothSame$gene_name, 
                             .tf = gene_bothSame$deletion,
                             .parents_or_hybrid = "hybrid") +
  ggtitle(paste("Both, same direction\n", gene_bothSame$gene_name, gene_bothSame$deletion, gene_bothSame$timepoint))
p_bothDifferent <- plotGenesTFdel(.gene_idxs = gene_bothDifferent$gene_name, 
                             .tf = gene_bothDifferent$deletion,
                             .parents_or_hybrid = "hybrid") +
  ggtitle(paste("Both, opposite directions\n", gene_bothDifferent$gene_name, gene_bothDifferent$deletion, gene_bothDifferent$timepoint))
ggarrange(p_Hyc, p_Hyp, p_bothSame, p_bothDifferent,
          nrow = 2, ncol = 2, common.legend = TRUE)
```
For the paper figure:
```{r figure_exampleGenes}
# # Hyc
# pdf("../Umich_Dissertation/chapters/figures/Networks/geneExampleHyc.pdf",
#     width = 3, height = 2.5)
# p_Hyc
# dev.off()
# # Hyp
# pdf("../Umich_Dissertation/chapters/figures/Networks/geneExampleHyp.pdf",
#     width = 3, height = 2.5)
# p_Hyp
# dev.off()
# # both same direction
# pdf("../Umich_Dissertation/chapters/figures/Networks/geneExampleBothSame.pdf",
#     width = 3, height = 2.5)
# p_bothSame
# dev.off()
# # both different direction
# pdf("../Umich_Dissertation/chapters/figures/Networks/geneExampleBothDifferent.pdf",
#     width = 3, height = 2.5)
# p_bothDifferent
# dev.off()
```


Scatter plots of log2 fold changes (with example genes labeled)

```{r hybrid allele lfc scatterplots}
p_both <- ggplot(data = filter(plotdf, padj_cer < p_thresh & padj_par < p_thresh), 
       aes(x = lfc_cer, y = lfc_par)) + 
  geom_point(aes(color = sign(lfc_cer) == sign(lfc_par))) +
  # # comment/uncomment second set of points to know where example
  # # genes are (ggMarginal can't handle second set of points though)
  # geom_point(data = bind_rows(gene_bothSame, gene_bothDifferent),
  #            aes(x = lfc_cer, y = lfc_par), 
  #            size = 3, shape = 1, color = "gold") +
  scale_color_discrete(limits = c(TRUE, FALSE),
                       type = c("black", "red")) +
  theme_classic() +
  theme(legend.position = "left") +
  ggtitle("genes with a significant effect\n in both hybrid alleles") +
  labs(color = "Fold changes are\nin the same direction",
       x = "Log2 fold change in Hyc",
       y = "Log2 fold change in Hyp") +
  geom_abline(intercept = 0, slope = 1)
plotdf_one <- filter(plotdf, !(padj_cer < p_thresh & padj_par < p_thresh)) |>
  slice_sample(n = 10000) |> 
  mutate(sig_species = if_else(padj_cer < p_thresh,
                               true = "Hyc",
                               false = "Hyp"))
p_one <- ggplot(plotdf_one, aes(x = lfc_cer, y = lfc_par)) + 
  geom_point(aes(color = sig_species), alpha = 0.25) +
  # # comment/uncomment second set of points to know where example
  # # genes are (ggMarginal can't handle second set of points though)
  # geom_point(data = bind_rows(gene_Hyc, gene_Hyp),
  #            aes(x = lfc_cer, y = lfc_par),
  #            shape = 1, size = 3, color = "gold") +
  scale_color_discrete(limits = c("Hyc", "Hyp"),
                       type = c(conservation_colordf$value[conservation_colordf$limits == "Scer"], 
                                conservation_colordf$value[conservation_colordf$limits == "Spar"])) +
  theme_classic() +
  theme(legend.position = "left") +
  ggtitle("genes with a significant effect\n in one hybrid allele") +
  labs(color = "Differentially\nexpressed allele",
       x = "Log2 fold change in Hyc",
       y = "Log2 fold change in Hyp") +
  geom_hline(yintercept = 0) +
  geom_vline(xintercept = 0)
p_both
p_one
```

For the paper figure (versions without ggMarginal so that we can highlight sample points):
```{r figure_both}
# pdf("../Umich_Dissertation/chapters/figures/Networks/hybridLFC_both_ExampleAnnotated.pdf", width = 5.5, height = 4)
# p_both
# dev.off()
# pdf("../Umich_Dissertation/chapters/figures/Networks/hybridLFC_one_ExampleAnnotated.pdf", width = 5.5, height = 4)
# p_one
# dev.off()
```
For the paper figure:
```{r figure_HybridLFCscatterplots}
pdf("../Umich_Dissertation/chapters/figures/Networks/hybridLFC_both.pdf",
    width = 5.5, height = 4)
grid::grid.draw(ggMarginal(p_both, groupColour = TRUE, groupFill = TRUE, margin = "both"))
dev.off()
pdf("../Umich_Dissertation/chapters/figures/Networks/hybridLFC_one.pdf",
    width = 5.5, height = 4)
grid::grid.draw(ggMarginal(p_one, groupColour = TRUE, groupFill = TRUE, margin = "both"))
dev.off()
```



```{r plotBarplot, warning=FALSE}
plotBarplot <- function(.updf, .downdf, .title) {
  p <- ggplot() +
    geom_histogram(data = .updf, 
                   aes(x = factor(deletion, 
                                  levels = TFnames),
                       y = after_stat(count),
                       fill = deletion), stat = "count") +
    geom_histogram(data = .downdf, 
                   aes(x = factor(deletion, 
                                  levels = TFnames),
                       y = -after_stat(count),
                       fill = deletion), stat = "count") +
    geom_hline(yintercept = 0) +
    scale_x_discrete(limits = TFnames,
                     breaks = TFnames) +
    theme_classic() +
    theme(legend.position = "none", axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) +
    xlab("") +
    ylab("number of DE genes") +
    ggtitle(.title)
  return(p)
}
plotdf <- sigdf |> 
  mutate(lfc_sign = sign(lfc)) |> 
  select(gene_name, deletion, organism, lfc_sign) |>
  group_by(gene_name, deletion) |> 
  summarise(orgs = list(organism),
            lfc_sign = unique(lfc_sign))

checkOrgs <- function(.query, .orgs, .excludeOthers = FALSE) {
  if (!.excludeOthers) {
    return(all(.query %in% unlist(.orgs)))
  }
  if (.excludeOthers) {
    #cat("performing setequal on\n", unique(unlist(.query)), "\n and\n",
        #unique(unlist(.orgs)), "\nwith lengths", length(.query), "and", length(.orgs))
    return(setequal(x = c(unlist(.query)), y = c(unlist(.orgs))))
  }
}
# tests for checkOrgs
test <- filter(plotdf, checkOrgs(.query = list("cer", "par"),
                                 .orgs = orgs,
                                 .excludeOthers = FALSE))
test |> filter(gene_name == "YMR242C" & deletion == "BAS1" & lfc_sign == -1) # should be present
table(unlist(test$orgs)) # can have hyb
test <- filter(plotdf, checkOrgs(.query = list("cer", "par"),
                                 .orgs = orgs,
                                 .excludeOthers = TRUE)) |> 
  select(gene_name, deletion) |> unique()
table(unlist(test$orgs)) # shouldn't have any hyb
nrow(test) # should have same nrow as Scer-Spar exclusive ring of Venn

# nEffects that are shared between all three yeasts:
filter(plotdf, checkOrgs(.query = list("cer", "hyb", "par"), 
                         .orgs = orgs, .excludeOthers = TRUE)) |> 
  select(gene_name, deletion) |> unique() # should be same nrow as Venn center count
```

```{r barplots, warning=FALSE}
max_nDE <- sigdf |> mutate(lfc_sign = sign(lfc)) |> 
  select(deletion, organism, gene_name, lfc_sign) |> 
  unique() |> group_by(deletion, organism, lfc_sign) |> summarise(n = n()) |>
  select(n) |> pull() |> max()
ylims_unique <- c(-max_nDE+10, max_nDE-10)
ylims_shared <- c(-100, 100)
# Scer unique
plotdf_cer <- filter(plotdf, checkOrgs(.query = list("cer"),
                                       .orgs = orgs,
                                       .excludeOthers = TRUE))
plotdf_cer_up <- filter(plotdf_cer, lfc_sign == 1)
plotdf_cer_down <- filter(plotdf_cer, lfc_sign == -1)
plotBarplot(.updf = plotdf_cer_up,
            .downdf = plotdf_cer_down,
            .title = "S. cer unique") +
  ylim(ylims_unique)

# Spar unique
plotdf_par <- filter(plotdf, checkOrgs(.query = list("par"),
                                       .orgs = orgs,
                                       .excludeOthers = TRUE))
plotdf_par_up <- filter(plotdf_par, lfc_sign == 1)
plotdf_par_down <- filter(plotdf_par, lfc_sign == -1)
plotBarplot(.updf = plotdf_par_up,
            .downdf = plotdf_par_down,
            .title = "S. par unique") +
  ylim(ylims_unique)

# Hyb unique
plotdf_hyb <- filter(plotdf, checkOrgs(.query = list("hyb"),
                                       .orgs = orgs,
                                       .excludeOthers = TRUE))
plotdf_hyb_up <- filter(plotdf_hyb, lfc_sign == 1)
plotdf_hyb_down <- filter(plotdf_hyb, lfc_sign == -1)
plotBarplot(.updf = plotdf_hyb_up,
            .downdf = plotdf_hyb_down,
            .title = "Hybird unique") +
  ylim(ylims_unique)

# Scer + Spar shared
plotdf_cerpar <- filter(plotdf, checkOrgs(.query = list("cer", "par"),
                                       .orgs = orgs,
                                       .excludeOthers = TRUE))
plotdf_cerpar_up <- filter(plotdf_cerpar, lfc_sign == 1)
plotdf_cerpar_down <- filter(plotdf_cerpar, lfc_sign == -1)
plotBarplot(.updf = plotdf_cerpar_up,
            .downdf = plotdf_cerpar_down,
            .title = "S. cer, S. par shared effects") +
  ylim(ylims_unique)

# Scer + Hyb shared
plotdf_cerhyb <- filter(plotdf, checkOrgs(.query = list("cer", "hyb"),
                                       .orgs = orgs,
                                       .excludeOthers = TRUE))
plotdf_cerhyb_up <- filter(plotdf_cerhyb, lfc_sign == 1)
plotdf_cerhyb_down <- filter(plotdf_cerhyb, lfc_sign == -1)
plotBarplot(.updf = plotdf_cerhyb_up,
            .downdf = plotdf_cerhyb_down,
            .title = "S. cer, Hybrid shared effects") +
  ylim(ylims_unique)

# Spar + Hyb shared
plotdf_parhyb <- filter(plotdf, checkOrgs(.query = list("par", "hyb"),
                                       .orgs = orgs,
                                       .excludeOthers = TRUE))
plotdf_parhyb_up <- filter(plotdf_parhyb, lfc_sign == 1)
plotdf_parhyb_down <- filter(plotdf_parhyb, lfc_sign == -1)
plotBarplot(.updf = plotdf_parhyb_up,
            .downdf = plotdf_parhyb_down,
            .title = "S. par, Hybrid shared effects") +
  ylim(ylims_unique)

# All 3 shared
plotdf_parhyb <- filter(plotdf, checkOrgs(.query = list("cer", "par", "hyb"),
                                       .orgs = orgs,
                                       .excludeOthers = TRUE))
plotdf_parhyb_up <- filter(plotdf_parhyb, lfc_sign == 1)
plotdf_parhyb_down <- filter(plotdf_parhyb, lfc_sign == -1)
plotBarplot(.updf = plotdf_parhyb_up,
            .downdf = plotdf_parhyb_down,
            .title = "S.cer, S. par, and Hybrid shared effects") +
  ylim(ylims_unique)
```
Sanity check: all the HAPS have more negative lfcs in hybrid-unique effects and hyb-par effects?
```{r}
sigdf |> filter(grepl("HAP", deletion)) |> mutate(lfc_sign = sign(lfc)) |> 
  select(organism, lfc_sign) |> table()
```

Second sanity check: MET28 has the most fully conserved
```{r}
sigdf |> 
  mutate(lfc_sign = sign(lfc)) |> 
  select(gene_name, deletion, organism, lfc_sign) |>
  group_by(gene_name, deletion) |> 
  summarise(orgs = list(organism),
            lfc_sign = unique(lfc_sign)) |> 
  ungroup() |> 
  filter(all(list("cer", "par", "hyb") %in% unlist(orgs))) |> 
  select(deletion) |> pull() |> table()
```
Third sanity check: BAS1/CBF1 have no hybrid-unique effects
```{r}
test <- sigdf |> 
  mutate(lfc_sign = sign(lfc)) |> 
  select(gene_name, deletion, organism) |>
  group_by(gene_name, deletion) |> 
  summarise(orgs = list(organism)) |> 
  ungroup() |> 
  filter(deletion %in% c("CBF1", "BAS1")) |> 
  filter(checkOrgs(.query = list("cer", "par"),
                     .orgs = orgs, .excludeOthers = FALSE)) |> 
  select(orgs) |> pull()
map(test, \(o) {
  return(paste0(sort(unique(unlist(o)))))
}) |> unlist() |> table()
```
How many replicates do BAS1/CBF1 have in hybrids?
```{r}
sample_info |> filter(genotype %in% c("BAS1delete", "CBF1delete") & organism == "hyb" & allele == "hyb") |> select(time_point_str, genotype) |> table()
```
1 in CBF1, that would do it.

## What percent of divergent vs conserved wildtype nitrogen responses are responding to TF deletions?
```{r}
sigdf <- TFdeldf |> filter(padj < p_thresh & abs(lfc) > eff_thresh)
genedf$sig <- genedf$gene_name %in% sigdf$gene_name
table(genedf$sig, genedf$is_conserved_unresponsive)
```


```{r message=FALSE}
# # tables/heatmaps for TP1 and TP3 of counts of how many genes are affected by each TFdel (either direction) in each species and hybrid alleles
# 
# effectsdf <- TFdeldf |> 
#   filter(timepoint != "TP2") |> 
#   group_by(deletion, organism, timepoint) |> 
#   summarise(nGenes = sum(abs(lfc) > eff_thresh & 
#                            padj < p_thresh)) |> 
#   pivot_wider(id_cols = c("deletion"), 
#               names_from = c("organism", "timepoint"), 
#               values_from = "nGenes") |> 
#   ungroup()
# effects_mat <- select(effectsdf, -deletion) |> as.matrix()
# rownames(effects_mat) <- effectsdf$deletion
# 
# # # In case you're curious if there's a relationship to differential expression of TFs between species (there isn't):
# # # ordering by expression difference (lfc) of TFs between species
# # tf_order <- TFdf |> filter(common %in% goodTFs & experiment == "LowN") |> 
# #   arrange(desc(effect_size_species)) |> select(common) |> pull()
# # tf_order <- c(tf_order, setdiff(goodTFs, tf_order))
# # # or mean expression level (also not related):
# # getMeanExpr <- function(.gene_idx) {
# #   return(mean(counts[.gene_idx,]))
# # }
# # tf_order <-  TFdf |> filter(common %in% setdiff(goodTFs, "HAP1")) |> 
# #   select(common, gene_name) |> unique()
# # 
# # tf_order$mean_expr <- map(tf_order$gene_name, getMeanExpr) |> 
# #   unlist() 
# # tf_order <- arrange(tf_order, desc(mean_expr)) |> select(common) |> pull()
# # tf_order <- c(tf_order, "HAP1")
# col_fun <- colorRamp2(c(0, 50, 300), c("blue", "yellow", "red"))
# p <- Heatmap(effects_mat, col = col_fun, na_col = "grey80",
#              column_order = c("cer_cer_TP1", "par_par_TP1", "cer_cer_TP3", "par_par_TP3", 
#                               "hyb_cer_TP1", "hyb_par_TP1", "hyb_cer_TP3", "hyb_par_TP3"),
#              # row_order = tf_order,
#              cell_fun = function(j, i, x, y, width, height, fill) {
#                output <- if_else(!(is.na(effects_mat[i, j])), 
#                                  true = as.character(effects_mat[i, j]), 
#                                  false = "-")
#                grid.text(output, x, y, gp = gpar(fontsize = 10))
#              })
# p
# 
# # # version including TP2
# # pdf("../../aligning_the_molecular_phenotype/paper_figures/TFdel/TFdel_heatmap_all3tps.pdf", 
# #     width = 5, height = 8)
# # Heatmap(effects_mat, col = col_fun, na_col = "grey80",
# #         column_order = c("cer_TP1", "par_TP1", "cer_TP2", "par_TP2", "cer_TP3", "par_TP3", 
# #                          "hyc_TP1", "hyp_TP1", "hyc_TP2", "hyp_TP2", "hyc_TP3", "hyp_TP3"),
# #         # row_order = tf_order,
# #         cell_fun = function(j, i, x, y, width, height, fill) {
# #           output <- if_else(!(is.na(effects_mat[i, j])), 
# #                             true = as.character(effects_mat[i, j]), 
# #                             false = "-")
# #           grid.text(output, x, y, gp = gpar(fontsize = 10))
# #         })
# # dev.off()
# 
# ### Repeating with sham, all TFdels were 2 randomly sampled WT replicates
# effectsdf <- TFdeldf_sham1 |> # change to sham2 to see if random sampling matters
#   filter(timepoint != "TP2") |> 
#   group_by(deletion, organism, timepoint) |> 
#   summarise(nGenes = sum(abs(lfc) > eff_thresh & 
#                            padj < p_thresh)) |> 
#   pivot_wider(id_cols = c("deletion"), 
#               names_from = c("organism", "timepoint"), 
#               values_from = "nGenes") |> 
#   ungroup()
# effects_mat <- select(effectsdf, -deletion) |> as.matrix()
# rownames(effects_mat) <- effectsdf$deletion
# 
# col_fun <- colorRamp2(c(0, 50, 300), c("blue", "yellow", "red"))
# p <- Heatmap(effects_mat, col = col_fun, na_col = "grey80",
#              column_order = c("cer_cer_TP1", "par_par_TP1", "cer_cer_TP3", "par_par_TP3", 
#                               "hyb_cer_TP1", "hyb_par_TP1", "hyb_cer_TP3", "hyb_par_TP3"),
#              # row_order = tf_order,
#              cell_fun = function(j, i, x, y, width, height, fill) {
#                output <- if_else(!(is.na(effects_mat[i, j])), 
#                                  true = as.character(effects_mat[i, j]), 
#                                  false = "-")
#                grid.text(output, x, y, gp = gpar(fontsize = 10))
#              })
# p
# ```
# 
# ## Volcano plots to compare power bwtn species
# ```{r}
# # basic volcano plot to compare power
# makeVolcanoPlot <- function(.tfdeldf, .tf, .org, .timepoint) {
#   plotdf <- .tfdeldf |> filter(organism == .org &
#                                  timepoint == .timepoint &
#                                  deletion == .tf)
#   ggplot(plotdf, aes(x = lfc, y = -log10(pval))) +
#     geom_point(aes(color = padj < p_thresh)) +
#     ylim(c(0, 15)) +
#     xlim(c(-5, 5))
# }
# 
# # At each TP, compare all 4 species with volcano plots
# # One example where Spar has flat pvals:
# makeVolcanoPlot(.tfdeldf = TFdeldf, .tf = "YAP1", .org = "cer", .timepoint = "TP3")
# #makeVolcanoPlot(.tfdeldf = TFdeldf, .tf = "YAP1", .org = "par", .timepoint = "TP3")
# # One example where Scer does:
# #makeVolcanoPlot(.tfdeldf = TFdeldf, .tf = "MET28", .org = "cer", .timepoint = "TP1")
# #makeVolcanoPlot(.tfdeldf = TFdeldf, .tf = "MET28", .org = "par", .timepoint = "TP1")
# # Example where Spar has a ton of TFdel replicates:
# #makeVolcanoPlot(.tfdeldf = TFdeldf, .tf = "HAP5", .org = "cer", .timepoint = "TP1")
# #makeVolcanoPlot(.tfdeldf = TFdeldf, .tf = "HAP5", .org = "par", .timepoint = "TP1")
# #makeVolcanoPlot(.tfdeldf = TFdeldf, .tf = "HAP5", .org = "cer", .timepoint = "TP3")
# #makeVolcanoPlot(.tfdeldf = TFdeldf, .tf = "HAP5", .org = "par", .timepoint = "TP3")
# # Spar has more detected, but Scer does also have a handful of sig (mainly TP3)
# ```
# 
# ## Deep dive: URE2, GAT1, and GLN3
# ```{r}
# makeUpsetPlot(.df = filter(TFdeldf, organism == "cer"),
#               .group_name = "deletion", 
#               .group_members = c("GLN3", "GAT1", "URE2"),
#               .item_names = "gene_name")
# # also true in par, hyc, hyp, but GAT1 and URE2 just have way fewer single effects and GLN3 has way more:
# makeUpsetPlot(.df = filter(TFdeldf, organism == "par"),
#               .group_name = "deletion", 
#               .group_members = c("GLN3", "GAT1", "URE2"),
#               .item_names = "gene_name")
# makeUpsetPlot(.df = filter(TFdeldf, organism == "hyc"),
#               .group_name = "deletion", 
#               .group_members = c("GLN3", "GAT1", "URE2"),
#               .item_names = "gene_name")
# makeUpsetPlot(.df = filter(TFdeldf, organism == "hyp"),
#               .group_name = "deletion", 
#               .group_members = c("GLN3", "GAT1", "URE2"),
#               .item_names = "gene_name")
# # so Spar's URE2-GAT1-GLN3 environment is dominant in the hybrid. 
# # Scer has much higher GAT1 expression, but Spar's GLN3 deletion has more effect
# # Other big difference: Scer has about 50 GAT1-URE2 shared genes, where Spar and hybrid only have 3-5
# # What are these genes?
# cer_tab <- TFdeldf |> filter(deletion %in% c("GAT1", "URE2") & 
#                      organism == "cer") |> select(gene_name) |> 
#   table() |> sort(decreasing = TRUE)
# cer_genes <- rownames(cer_tab)[cer_tab == 2]
# par_tab <- TFdeldf |> filter(deletion %in% c("GAT1", "URE2") & 
#                                 organism == "par") |> select(gene_name) |> 
#   table() |> sort(decreasing = TRUE)
# par_genes <- rownames(par_tab)[par_tab == 2]
# intersect(cer_genes, par_genes) # DUR12, MEP1, both urea/ammonia enzymes
# # where are they in non-Scer organisms?
# TFdeldf |> filter(gene_name %in% setdiff(cer_genes, par_genes)) |> 
#   select(deletion, organism) |> table()
# # two conclusions: 
# # 1) either both parents have a lot of effects or both hybrid alleles do (GLN3, GCR2, MET28 are exceptions where all 4 have many effects)
# # 2) when there is a stark difference in which TF affects which organism, the Scer parental gene copies are most sensitive (SOK2, STB5, RTG3, URE2, GZF3, GAT1, GCN4, ARG81)
# 
# # what clusters are these genes in?
# 
# # TODO: b/c we're not doing clustering for this TFdel paper, change to "conserved decreasing" and "conserved increasing" (and maybe a third "uncorrelated" category if you want to use it? Only if it's helpful to provide biological insight with the DALs)
# 
# # finaldf |> filter(gene_name %in% setdiff(cer_genes, par_genes)) |> 
# #   select(cer, par) |> table() # heavily conserved, but even split between 1-1 and 2-2
# # gene_idxs11 <- finaldf |> filter(gene_name %in% setdiff(cer_genes, par_genes) &
# #                                    cer == 1 & par == 1) |> select(gene_name) |> pull()
# # gene_idxs22 <- finaldf |> filter(gene_name %in% setdiff(cer_genes, par_genes) &
# #                                    cer == 2 & par == 2) |> select(gene_name) |> pull()
# # getGOSlimDf(.idxs = gene_idxs11, .group_name = "GAT1_URE2_ScerUnique11")
# # Conserved increasing: 12 catabolic genes (including allatonin degredation genes DALs 1, 2, 4, 7)
# # getGOSlimDf(.idxs = gene_idxs22, .group_name = "GAT1_URE2_ScerUnique22") 
# # Conserved decreasing: 18 nucleolus/15 rRNA processing genes, 6 helicase (mainly DEAD box RNA helicases for rRNA processing)
# 
# # # What timepoint is the dynamics affected? If this is related to the URE2-GAT1 interaction, we'd expect the LowN timepoint
# # # 1-1
# # plotGenesTFdel(.gene_idxs = gene_idxs11, .tf = "GAT1") # yes later timepoint, mainly Scer but a little Spar
# # plotGenesTFdel(.gene_idxs = gene_idxs11, .tf = "URE2") # both timepoints, just Scer
# # # what about hybrid? 
# # plotGenesTFdel(.gene_idxs = gene_idxs11, .tf = "GAT1", .parents_or_hybrid = "hybrid")
# # plotGenesTFdel(.gene_idxs = gene_idxs11, .tf = "URE2", .parents_or_hybrid = "hybrid")
# # # Interestingly there is an effect at the early timepoint in both hybrid alleles. Why weren't they flagged as dynamic effects?
# # effectsdf |> filter(organism %in% c("hyc", "hyp") &
# #                      deletion %in% c("GAT1", "URE2") &
# #                      gene_name %in% gene_idxs11) |> select(effect, deletion) |> table()
# # # They're all URE2 only, but there is a slight effect in GAT1
# # TFdeldf |> filter(organism %in% c("hyc", "hyp") &
# #                     deletion == "GAT1" &
# #                     timepoint == "TP1" &
# #                     gene_name %in% gene_idxs11) |> 
# #   arrange(desc(lfc)) # Only enough power to detect a couple, namely the very classy haze protective factor HPF1, which reduces cloudiness in white wines
# # # But it's not just HPF1/YOL155C causing this -- most hybrid orthologs are up decently in URE2, just without significant pvalues
# # 
# # # 2-2
# # plotGenesTFdel(.gene_idxs = gene_idxs22, .tf = "GAT1") # yowza later timepoint, only Scer
# # plotGenesTFdel(.gene_idxs = gene_idxs22, .tf = "URE2") # mostly later timepoint, also yowza, only Scer
# # # hybrid
# # plotGenesTFdel(.gene_idxs = gene_idxs22, .tf = "GAT1", .parents_or_hybrid = "hybrid")
# # plotGenesTFdel(.gene_idxs = gene_idxs22, .tf = "URE2", .parents_or_hybrid = "hybrid")
# # # Here the hybrid effect at TP3 is in the opposite direction (and weaker)
# # 
# # # Taking stock: identified ~50 genes that had dynamics affected by both GAT1 and URE2 deletion
# # # in Scer but not in Spar or hybrid alleles
# # # Interesting b/c GAT1 is expressed much higher in Scer (and hyc allele transiently in response to LowN)
# # # These genes are 50-50 split between increasing and decreasing cluster, but they have all conserved dynamics btwn species
# # # GAT1/URE2 deletion affects late timepoint the strongest in Scer, no effect in Spar
# # # in hybrid, there is some evidence of an early timepoint effect---especially URE2 delete on increasing genes
# # # but the late timepoint effect is really only seen in Scer
# # 
# # # investigating GAT1 Scer_only regulated genes
# # # first increasing genes
# # gene_idxs_par <- getGeneGroup(.tf = "GAT1", .tp = "TP3", .lfc_sign = 1,
# #                               .clust = 2, .org = "par")
# # gene_idxs_cer <- getGeneGroup(.tf = "GAT1", .tp = "TP3", .lfc_sign = 1,
# #                               .clust = 2, .org = "cer")
# # plotdf <- getGroupDf(.tf = "GAT1", .tp = "TP3", .lfc_sign = 1,
# #                      .clust = 2, .genes_cer = gene_idxs_cer,
# #                      .genes_par = gene_idxs_par)
# # ggplot(plotdf, aes(x = lfc_cer, y = lfc_par)) + 
# #   geom_point(aes(color = tf_effect)) +
# #   geom_abline(slope = 1, intercept = 0)
# # # Scer-only TF response, conserved dynamics
# # gene_idxs <- plotdf |> filter(tf_effect == "Scer only" & dynamics == "conserved" &
# #                                 par == 2) |> 
# #   select(gene_name) |> pull()
# # plotGenesTFdel(.tf = "GAT1", .gene_idxs = gene_idxs)
# # plotGenesTFdel(.tf = "GAT1", .gene_idxs = sample(gene_idxs, size = 12), .single_genes = TRUE, .normalization = "scale")
# # # (there are no decreasing genes for this cluster)
# # 
# # # investigating GAT1 Scer_only regulated genes
# # # first increasing genes
# # gene_idxs_par <- getGeneGroup(.tf = "GAT1", .tp = "TP3", .lfc_sign = 1,
# #                               .clust = 1, .org = "par")
# # gene_idxs_cer <- getGeneGroup(.tf = "GAT1", .tp = "TP3", .lfc_sign = 1,
# #                               .clust = 1, .org = "cer")
# # plotdf <- getGroupDf(.tf = "GAT1", .tp = "TP3", .lfc_sign = 1,
# #                      .clust = 1, .genes_cer = gene_idxs_cer,
# #                      .genes_par = gene_idxs_par)
# # ggplot(plotdf, aes(x = lfc_cer, y = lfc_par)) + 
# #   geom_point(aes(color = tf_effect)) +
# #   geom_abline(slope = 1, intercept = 0)
# # # Scer-only TF response, conserved dynamics
# # gene_idxs <- plotdf |> filter(tf_effect == "Scer only" & dynamics == "conserved" &
# #                                 par == 1) |> 
# #   select(gene_name) |> pull()
# # plotGenesTFdel(.tf = "GAT1", .gene_idxs = gene_idxs)
# # plotGenesTFdel(.tf = "GAT1", .gene_idxs = sample(gene_idxs, size = 12), .single_genes = TRUE, .normalization = "scale")
# # # next decreasing genes
# # gene_idxs_par <- getGeneGroup(.tf = "GAT1", .tp = "TP3", .lfc_sign = -1,
# #                               .clust = 1, .org = "par")
# # gene_idxs_cer <- getGeneGroup(.tf = "GAT1", .tp = "TP3", .lfc_sign = -1,
# #                               .clust = 1, .org = "cer")
# # plotdf <- getGroupDf(.tf = "GAT1", .tp = "TP3", .lfc_sign = -1,
# #                      .clust = 1, .genes_cer = gene_idxs_cer,
# #                      .genes_par = gene_idxs_par)
# # ggplot(plotdf, aes(x = lfc_cer, y = lfc_par)) + 
# #   geom_point(aes(color = tf_effect)) +
# #   geom_abline(slope = 1, intercept = 0)
# # # Scer-only TF response, conserved dynamics
# # gene_idxs <- plotdf |> filter(tf_effect == "Scer only" & dynamics == "conserved" &
# #                                 par == 1) |> 
# #   select(gene_name) |> pull()
# # plotGenesTFdel(.tf = "GAT1", .gene_idxs = gene_idxs)
# # plotGenesTFdel(.tf = "GAT1", .gene_idxs = sample(gene_idxs, size = 12), .single_genes = TRUE, .normalization = "scale")
```
